Retirement Consumption: Evidence from UK Pension Reform

### 2023-07-27 Texfiles/Draft1.tex
Potential finish

        Word count: [Insert]
        \vfill
        I confirm that this is entirely my own work and has not previously been submitted for assessment, and I have read and understood the University’s and Faculty’s definition of Plagiarism
    \end{center}
\end{titlepage}
\newpage
\maketitle
\begin{abstract}
    I find the consumption response of early retirees to the `pensions freedom
    act', which meant defined contribution pensions could be used for other
    options than just purchasing an annuity. I use a discontinuity in year of
    retirement and a difference in difference approach using those who were
    unaffected by the policy reform as a natural control group. I then solve
    three sets of lifecycle models from age of retirement until an agent reaches
    110. I simulate their consumption with and without annuities, i.e. a the
    policy reform, and match their consumption response to what was seen in the
    empirical models. I find that a lifecycle model with a bequest motive best
    fits this consumption response.
\end{abstract}
\newpage
\tableofcontents
\newpage

\section{Introduction}
Classical economic theory suggests that retirement annuities, which provide a
guaranteed income stream until the end of life, should be highly valued by
individuals as a way to insure against late death \citep{yaari_65}. However, in
developed countries, rates of annuitisation are far below the levels that theory
predicts -- this phenomenon has been termed the ``annuity puzzle". The literature
has suggested several reasons for this, but there is no consensus about which
mechanism is dominant. I exploit early retirees' consumption responses to a
reform to annuity policy in the UK to add to this literature. [\textbf{Add
            something once I have a conclusion}].

In the UK, adults have three methods of funding retirement: the state pension;
defined benefit (DB) employee pension schemes, which provide an income in
retirement based on years of service and a function of wages; and defined
contribution (DC) pensions, under which individuals save and invest in a
tax-advantaged account that is usually supplemented with employer contributions.
Under the 2010-2015 coalition government in the UK, the law regarding the use of
private DC pensions at retirement changed. Individuals were no longer forced to
annuitise their pension pots and could access them in a variety of ways, such as
a lump sum withdrawal or income drawdown, involving a steady withdrawal of
assets from the pension pot -- common advice is to take 4\% a year -- whilst the
rest remains invested. Subsequently, the number of annuities sold in the UK
dropped precipitously.

@@

\cite{yaari_65} showed that under standard assumptions with risk averse agents
we would expect individuals to annuitise all of their wealth at retirement to
insure against the risk of long life. And more recently
\cite{davidoff_brown_diamond_aer_2005} show that complete annuitisation is
optimal under a broader range of assumptions. Since then there has been much
discussion of possible reasons for why people do not annuitise.
\cite{finkelstein_porteba_2002, finkelstein_porteba_2004} find evidence of
adverse selection into the UK annuity market and \cite{mitchell_et_al_1999} find
evidence of the same in the US. This means longer lived individuals select into
buying annuities which drives up annuity prices because providers need to pay
individuals for longer than they originally thought they would need to. This
makes the `money's worth' of an annuity lower for the general population as
opposed to the population of annuitants, putting off individuals from buying an
annuity. However, \cite{mitchell_et_al_1999} also find that theory would still
predict annuitisation because the money's worth of annuities is not that far
from `actuarially fair' (this is the where the present discounted value of annuities
equals the price that an individual pays for it).

\cite{friedman_warshawsky_qje_1990} show that annuitisation decisions can be
fully explained by a mixture of bequest motives and `actuarially unfair'
annuities. They solve an augmented life-cycle model and calculate annuity demand
for a range of differences between rate of return assets would get in the stock
market and the rate of return annuities give an individual. For plausible values
they find that individuals would optimally not annuitise much wealth. Similarly
to \cite{finkelstein_porteba_2004}, \cite{friedman_warshawsky_chicago_1988} show
that there is a significant difference between the life expectancy of annuitants
and the general population in the American annuity market, which therefore
implies a degree of adverse selection. But they find this cannot fully explain
the annuitisation puzzle and only when bequest motives are added to the model
can annuitisation rates be rationalized.

More recently, \cite{lockwood_red_2012} builds on this and shows that a
realistic bequest motive in lifecycle simulations achieves realistic
annuitisation rates. He solves a simple lifecycle model with bequest motives
taken from several recent papers in the retirement savings literature. The
bequest motives he picks matches other important aspects of empirical
distribution of bequests such as how much individuals give and the financial
characteristics of these individuals. Also, \cite{odea_sturrock_rest_2023} find
that subjective life expectancies can explain a large part of the annuitisation
puzzle especially if agents are less risk averse.

Apart from bequests and annuity prices, a range of other reasons for the annuity
puzzle have been explored. Pre-cautionary savings for unexpected end-of-life
medical expenses might cause individuals to prefer holding assets rather than a
an annuity. Moreover, \cite{vidalmelia_lejarragagarcia_munich_2004} suggests
that couples might share the risk of living for a long-time between them and
therefore couples may be less likely to annuitise their wealth.

Given that uncertain end-of-life medical expenses are not such an issue in the
UK as opposed to the US, and that the focus of recent literature has been on the
bequest model and uncertain life expectancy, I choose also to focus on bequest
and objective life expectancy in the UK context.

@@

I now run the empirical models on this simulated data. This lets us
see which lifecycle model gives us the closest match to the coefficients seen in the
real data.

@@

The interaction models in columns 3 and 4 both have correct signs for the main
reform effect in the standard and bequest models, although in both the
interaction with financial wealth has a negative coefficient showing that the
increase in consumption caused by the reform was progressively lower for those
who would have had to annuitise more wealth. In the subjective model table the
signs are again wrong apart from for the main treatment effect in the financial
interaction column. The lifecycle model that appears closest to the empirical
results is again the bequest model, which matches the signs from Tables
\ref{tab:DcOnlyInteract} and \ref{tab:DcOnlyFinWealthInteract}, but the
coefficients are smaller.

Across all regressions, the lifecycle model using subjective life expectancies
does not match the change in consumption that is seen when using ELSA data. It
is surprising that the consumption response is also very different to the other
lifecycle models used. One possible explanation for this could be that the
control and treatment groups have quite different sets of objective life
expectancies which therefore cause different consumption paths in early
retirement. The structure that the Weibull distribution imposes on the
death probabilities could also be causing discontinuities in a given
individuals subjective probability of death between two periods, which would
impact how they consume across that part of retirement.

The evidence just presented points towards predictions from the bequest model
being best able to replicate the consumption response in the data. As
highlighted above, the benefit of having a theoretical model is that we can
trace out future values of consumption for individuals, only needing to know
their starting point. So using the bequest solution to the retirement problem, I
plot the whole path of average consumption and assets for the ELSA treated
cohort. This plot is shown in the figure below.

\textbf{Also get some plots of consumption in early retirement with and without
    annuities for the different model types}

@@

I have shown that consumption in early retirement increased overall and
particularly on

\subsection{Rough plan}
\begin{itemize}
    \item Intro
          \begin{enumerate}
              \item "Latest data from HM Revenue Customs (published in April)
                    showed more than £45bn has been taken from pots since 2015."
                    https://www.ftadviser.com/pensions/2022/02/08/pension-freedoms-were-they-really-a-good-idea/
              \item what portion of savings are because of retirement savings?
              \item Also talk about heterogeneity across countries. Some
                    countries want to move towards more annuitisation. This
                    annual review is a good source of info
                    \cite{banks_crawford_ar_2022}
              \item https://www.moneysavingexpert.com/savings/pension-freedom/
          \end{enumerate}

          \textbf{Follow up by changing chat because the parameter estimates for
              the diff in diff have changed a bit -- DONE but check again when I
              have gone through it -- haven't checked the comparisons }
    \item \textbf{Abstract}
    \item \textbf{Add bit about testing just those people who retired when they
              expected to retire.}
    \item \textbf{Could add consumption and retirement year plot?}
    \item \textbf{Maybe a small discussion of controls for the main regression
              tables as well}
    \item Conclusion
\end{itemize}

>>>>

        Word count: [Insert]
        \vfill
        I confirm that this is entirely my own work and has not previously been submitted for assessment, and I have read and understood the University’s and Faculty’s definition of Plagiarism
    \end{center}
\end{titlepage}
\newpage
\maketitle
\begin{abstract}
    I identify the consumption responses of retirees to the UK's `pensions
    freedom act', a landmark law change that meant defined contribution pension
    holders were no longer forced to annuitise their funds and could instead
    access their pensions in a variety of ways. I use a discontinuity in year of
    retirement and a difference-in-difference approach, harnessing those who
    were unaffected by the policy reform as a natural control group. I then
    solve three sets of lifecycle models, tracking assets from retirement until an
    agent reaches age 110. I simulate their consumption with and without
    annuities, mimicking the policy reform, and match their consumption
    responses to what was seen in the empirical models. I find that a lifecycle
    model with a bequest motive best fits the identified consumption response —
    a result that could inform policymakers in designing optimal pension
    policies for an ageing population.
\end{abstract}
\newpage
\tableofcontents
\newpage

\section{Introduction}
Classical economic theory suggests that retirement annuities, which provide a
guaranteed income stream until the end of life, should be highly valued by
individuals as a way to insure against late death \citep{yaari_65}. However, in
developed countries, rates of annuitisation are far below the levels that theory
predicts -- this phenomenon has been termed the ``annuity puzzle". The literature
has suggested several reasons for this, but there is no consensus about which
mechanism is dominant. I exploit early retirees' consumption responses to a
reform to annuity policy in the UK to add to this literature.

In the UK, adults have three methods of funding retirement: the state pension;
defined benefit (DB) employee pension schemes, which provide an income in
retirement based on years of service and wages; and defined
contribution (DC) pensions, under which individuals save and invest in a
tax-advantaged account that is usually supplemented with employer contributions.
Under the 2010-2015 coalition government in the UK, the law regarding the use of
private DC pensions at retirement changed. Individuals were no longer forced to
annuitise their pension pots and could access them in a variety of ways, such as
a lump sum withdrawal or income drawdown, involving a steady withdrawal of
assets from the pension pot -- common advice is to take 4\% a year -- whilst the
rest remains invested. Subsequently, the number of annuities sold in the UK
dropped precipitously.

@@

\cite{yaari_65} showed that under standard assumptions with risk averse agents
we would expect individuals to annuitise all of their wealth at retirement to
insure against the risk of long life. More recently,
\cite{davidoff_brown_diamond_aer_2005} show that complete annuitisation is
optimal under a broader range of assumptions. Since then, there has been much
discussion of possible reasons for why people choose not to annuitise.
\cite{finkelstein_porteba_2002, finkelstein_porteba_2004} find evidence of
adverse selection into the UK annuity market and \cite{mitchell_et_al_1999} find
evidence of the same in the US. This means longer-lived individuals select into
buying annuities which drives up annuity prices because providers need to pay
individuals for longer than they originally predicted they would need to. This
makes the `money's worth' of an annuity lower for the general population as
opposed to the population of annuitants, putting off individuals from buying an
annuity. However, \cite{mitchell_et_al_1999} also find that theory would still
predict annuitisation because the money's worth of annuities is not that far
from `actuarially fair' (meaning the present discounted value of annuities
equals the price that an individual pays for it).

\cite{friedman_warshawsky_qje_1990} show that annuitisation decisions can be
fully explained by a mixture of bequest motives and `actuarially unfair'
annuities. They solve an augmented lifecycle model and calculate annuity demand
for a range of differences between the rate of return assets would get in the stock
market and the rate of return annuities give an individual. For plausible values,
they find that individuals would optimally not annuitise much wealth. Similarly
to \cite{finkelstein_porteba_2004}, \cite{friedman_warshawsky_chicago_1988} show
that there is a significant difference between the life expectancy of annuitants
and the general population in the American annuity market, which therefore
implies a degree of adverse selection. But, they find this cannot fully explain
the annuitisation puzzle, and only when bequest motives are added to the model
can annuitisation rates be rationalised.

More recently, \cite{lockwood_red_2012} builds on this and shows that a
realistic bequest motive in lifecycle simulations achieves realistic
annuitisation rates. He solves a simple lifecycle model with bequest motives
taken from several recent papers in the retirement savings literature. The
bequest motives he picks match other important aspects of the empirical
distribution of bequests, such as how much individuals bequeath and the financial
characteristics of these individuals. Also, \cite{odea_sturrock_rest_2023} find
that subjective life expectancies can explain a large part of the annuitisation
puzzle, especially if agents are less risk averse.

Apart from bequests and annuity prices, a range of other reasons for the annuity
puzzle have been explored. Precautionary savings for unexpected end-of-life
medical expenses might cause individuals to prefer holding assets rather than
an annuity. Moreover, \cite{vidalmelia_lejarragagarcia_munich_2004} suggest
that couples might share the risk of long life by pooling their resources, and
therefore couples may be less likely to annuitise their wealth.

Given that uncertain end-of-life medical expenses are not such an issue in the
UK as opposed to the US, and that the focus of recent literature has been on the
bequest model and uncertain life expectancy, I choose also to focus on bequest
and objective life expectancy in the UK context. To my knowledge, there has been
no attempt to analyse the consumption response to 2014 reform, nor evaluate what
common lifecycle models predict for the policy change.

@@

I now run the empirical models on this simulated data. This lets us see which
lifecycle model gives us the closest match to the coefficients seen in the real
data. In the structural modelling literature, it is often common to run policy
tests on estimated models to see if simulated responses match real world
responses, for example see \cite{mcgee_2021}. Due to the time constraints of
this project, I did not estimate a full structural model but I do check which
lifecycle model fits the response seen in the ELSA data.

@@

The interaction models in columns 3 and 4 both have correct signs for the main
reform effect in the standard and bequest models, although in both the
interaction with financial wealth has a negative coefficient, showing that the
increase in consumption caused by the reform was progressively lower for those
who would have had to annuitise more wealth. In the subjective model table, the
signs are again wrong apart from for the main treatment effect in the financial
interaction column. The lifecycle model that appears closest to the empirical
results is again the bequest model, which matches the signs from Tables
\ref{tab:DcOnlyInteract} and \ref{tab:DcOnlyFinWealthInteract}, but the
coefficients are smaller.

Across all regressions, the lifecycle model using subjective life expectancies
does not match the change in consumption that is seen when using ELSA data. It
is surprising that the consumption response is also very different to the other
lifecycle models used. One possible explanation for this could be that the
control and treatment groups have quite different sets of subjective life
expectancies, which therefore cause different consumption paths in early
retirement. This heterogeneity in death probabilities would not occur in the
with the other models because the objective death probabilities at a given age
for different birth cohorts are not that large. The structure that the Weibull
distribution imposes on the death probabilities could also be causing
discontinuities in a given individuals subjective probability of death between
two periods, which would impact how they consume across that part of retirement.

The evidence just presented points towards simulated data using the bequest
model being best able to replicate the consumption response observed in the
empirical models. As highlighted above, the benefit of having a theoretical
model is that we can trace out the asset path of consumption for individuals,
only needing to know their retirement wealth.

@@

The results from the empirical section of this paper point toward there being
some impact of the `pensions freedom act' on the consumption of retirees early
in retirement. The specifications that used the full available data set show a
clear increase in total monthly consumption, and particularly an increase in food
consumption away from home and on clothes. Although, this increase of 36 a month
only equates to 432 a year, over a longer period it may mean some retirees end
up with fewer resources than they require toward the end of their life.

Moreover, the results from the lifecycle section show not just that the models
create different consumption predictions, but they also add to the evidence that
lifecycle models with bequests can do a good job of explaining the consumption
patterns of retirees. That simulated data from a bequest model matches the
consumption response also points towards bequests being able to explain the
annuitisation puzzle.

Since the bequest motive model fits the consumption response best, we can worry
about the increase in early retirement consumption less. This is because,
as highlighted in the analysis of the median individual, assets will diverge
from the objective and subjective lifecycle models during retirement, reducing
at a slower rate. Therefore, the government should not worry about spending
early in retirement being linked to low assets at the end of an individuals
life.

Future research is still needed to understand the full impact the reform has had
on consumption across the whole of retirement, especially as more data on
retirees impacted by the reform becomes available. Moreover, as the number of
individuals retiring with a defined contribution pension grows, understanding
the mechanisms behind their choices will matter more to policy makers. For
example, the advice that is provided to individuals on retirement.
